---
title: "Lecture 1"
output: html_notebook
author: Bob Stine
date: 27 Jun 2024
---

Contents:

    * Basic R commands
    * Examples of time series plots using astsa
    
Before running the following commands, become familiar with the different panes within the R-Studio window.

Next, use the Package pane to install the `astsa` package. We will use this package throughout the course.

Once other packages are installed that support output formatting (the software will bug you about this), R-Studio will create an html version of your R notebook and save the file adjacent to the notebook file (identified by the .Rmd filename suffix).  I seldom use these, but they can be helpful for sharing what you've done with someone that doesn't have R/R-Studio.


# - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -

# Basic R commands

# - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -

You can use R as a powerful calculator with many built-in functions.  

You can put multiple commands within the same input block for evaluation. R-Studio button/keyed commands let you evaluate one line or the entire block. The output of running a line (or "chunk" of several lines or commands) appears below.

```{r}
4 + 5

log(34)  # natural log (this is a comment)
```

Vectors (sequences) and assignment, with error messages and NaN/Infinity detection.

```{r}
x <- c(-1,0,1,2,3,4,5)   # assignment (you can use =, but I won't)
x
x + 3
log(x)
```

Indexing

```{r}
length(x)

x[0]   # R is 1-based indexing
x[1]
x[3]

x[45]
```

Matrices are a fundamental data type in R.  Matrices are constructed in a variety of ways, but most directly with the `matrix` function.

```{r}
m <- matrix(1:6, nrow=3, ncol=2)
m
```

```{r}
dim(m)
```

Another way to create a matrix is with the function `cbind`.

```{r}
cbind(1:3, 4:6)
```

Data frames are rectangular objects that allow different data types, such as text and numbers.

```{r}
data.frame(letter=c('a', 'b', 'c', 'd'), index=1:4)
```



# - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -

# Random numbers, stats, plots

# - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -

Random numbers, sequences, plots.

```{r}
set.seed(234)   # to replicate pseudo-random examples

n <- 50     
x <- 1:n        # watch out Python users
y <- rnorm(n)

plot(x,y)       # standard R scatter plot
```

Random walk

```{r}
w <- cumsum(y)
x[1:2]
w[1:2]
plot(x,w)
```

"Fancier" version.

```{r}
par(mgp = c(2, 1, 0), mar = c(3, 3, 2, 1))       # what do you think this does; ?par to find out
plot(x,w, xlab="Time", ylab=bquote(Random~Walk~W[t]), pch=4, col='dodgerblue')
lines(x,w, col='lightgray', lwd=0.5)
```

Histograms and distributions.

```{r}
mean(w)
sd(w)

hist(w, probability = TRUE)
lines(density(w), col=2, lwd=3)
```

Regression is built-in.

```{r}
summary(lm(w ~ x))    # should this be significant?
```


# - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -

# Time series in R

# - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -

The random walk sequence `w` created above is simply a sequence.  R doesn't treat it special; `w` is a vector of numbers with no special attributes.

```{r}
typeof(w)

attributes(w)
```

Time series objects in R have a time sequence attribute that influences the x-axis in plots in *some* functions.

```{r}
w.ts <- ts(w, start=2000)

typeof(w.ts)
attributes(w.ts)
```

The built-in function `plot` uses the ts attribute to determine the x-axis.  

```{r}
plot(w.ts)
```

Attributes are important for aligning time series when lagging, shifting a time series foreward or backward in time.  For example, one might want to fit a simple autoregression of the form
$$ \hat{w}_t = \hat\alpha + \hat\beta\,w_{t-1} $$

```{r}
w.ts[1:2]      # index as a sequence, not by the underlying ts indices shown in plot!

lag(w.ts,1)    # changes the underlying time series attribute
```

```{r}
lm(w.ts ~ lag(w.ts))    # not what we want!  What happened???
```

WARNING: Many functions in R don't pay attention to the time series attribute.  Fortunately, `cbind` does. Or, you can use `ts.intersect` to keep just the complete cases.

```{r}
head(  cbind(w=w.ts, lag_w=lag(w.ts))    )   # adding column names is optional, but useful
```
```{r}
head(   ts.intersect(w=w.ts, lag_w=lag(w.ts))    )
```

Now we can get the correct estimated autoregression.

```{r}
rwData = ts.intersect(w=w.ts, lag_w=lag(w.ts))     # why are there only 47 d.f?
summary(lm(w ~ lag_w, data=rwData))
```


# - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -

# ASTSA package

# - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -

Need to load the package into the R environment first. (You have to install the package first.)

```{r}
library(astsa)
```

The astsa package includes the function `tsplot` which makes nicer figures.

```{r}
?tsplot # or type the name into the help panel
```

```{r}
tsplot(w.ts)
```

ggplot is a popular alternative, albeit at the cost of learning a different approach to setting options. It has another weakness too: ggplot doesn't respect time series objects (too bad).

```{r}
library(ggplot2)
```

```{r}
df <- data.frame(time=1:length(w), w = w)
ggplot(df, aes(x=time, y=w)) + geom_line()
```

## Example 1.1   J&J earnings per share

The time series `jj` is included in the astsa package.

```{r}
?jj                # get an error if you don't have astsa loaded

attributes(jj)     # how can you tell it is quarterly data
```

```{r}
tsplot(jj)
```

```{r}
tsplot(log10(jj))
```

Why switch to natural logs?

```{r}
tsplot(log(jj), main="J&J Quarterly Earnings", ylab="Log Earnings")
abline( lm(log(jj) ~ time(jj)), col='red')
```

What's the interpretation of the fitted line?

```{r}
coefficients( lm(log(jj) ~ time(jj)) )
```
Darn...  The option to use a log scale with untransformed units on the y-axis doesn't play well with the added line.

```{r}
tsplot(jj, main="J&J Quarterly Earnings", ylab="Earnings", log='y')
abline( lm(log(jj) ~ time(jj)), col='red')
```


## Example 1.2   Global warming

Both series are temperature anomalies.

```{r}
?gtemp_ocean
?gtemp_land
```

Plot together or overlaid  The first plot shown determines the axes limits.

```{r}
layout(matrix(c(1,2,3,3), nrow=2, byrow=TRUE))

tsplot(gtemp_ocean, col='blue', ylab="Ocean Temp Dev")
tsplot(gtemp_land, col='brown', ylab="Land Temp Dev")

tsplot(gtemp_ocean, col='blue', ylab="Temp Deviation")  
lines(gtemp_land, col='brown')   # add lines to an existing plot
```

Cleaned up a bit with a legend in the plot.

```{r}
tsplot(gtemp_ocean, col='blue', ylim=c(-.8, 1.5), lwd=2)
lines(gtemp_land, col='brown')
legend('topleft', legend=c("Ocean", "Land"), lwd=2:1, col=c('blue', 'brown'))
```


## Example 1.6   fMRI

Response to input electrical signal.

```{r}
?fmri1
```

```{r}
attributes(fmri1)
```

```{r}
head(fmri1)
```

```{r}
dim(fmri1)
```

Create a vector for the stimulus signal to show in the plot.

```{r}
rep(1,4)

rep(1:3,2)
```

```{r}
signal <- rep( c(rep(0.6,16), rep(-0.6,16)), 4)
length(signal)
```

Clear response in the cortex. Less clear elsewhere. 

```{r}
nn <- colnames(fmri1)
j <- 4
tsplot(fmri1[,j  ], main=nn[j], ylim=c(-0.6,0.6), ylab="BOLD")
lines( fmri1[,j+1], col=4)
lines(signal, type='s', col=2)
```

Not so clear here.  We'll have a better way to detect this signal later in the course.

```{r}
j <- 6
tsplot(fmri1[,j  ], main=nn[j], ylim=c(-0.6,0.6))
lines( fmri1[,j+1], col=4)
lines(signal, type='s', col=2)
```

```{r}
j <- 8
tsplot(fmri1[,j  ], main=nn[j], ylim=c(-0.6,0.6))
lines( fmri1[,j+1], col=4)
lines(signal, type='s', col=2)
```

